Developing effective theories for single and bilayer graphene using density matrix downfolding

### 2019-07-30 draft.tex
edited + shortened draft. have 1.5 pages for references now.

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{subfigure}
\usepackage[margin=1.0in]{geometry}
\usepackage{float}
\usepackage{ulem}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{wrapfig}
\renewcommand{\baselinestretch}{1.5}
\usepackage[utf8]{inputenc}
\author{Shivesh Pathak}
\title{Developing effective theories for the square lattice Hubbard model using density matrix downfolding}

\begin{document}
\maketitle
\begin{abstract}
A core problem of condensed matter physics is understanding how to systematically develop low-energy effective models for strongly correlated electron systems. 
We believe that a density matrix downfolding (DMD) method can be used to systematically build effective theories for such systems beginning with a high energy theory like the \textit{ab-initio} Hamiltonian H$_\text{ab}$.
We present as a preliminary result an effective theory developed using DMD which can accurately reproduce the spectra and properties of the lowest lying excitations of the CuO molecule.
As a step towards bulk strongly correlated materials, we propose developing effective theories using DMD for the simplest toy system which exhibits strongly correlated behavior: the 1-band square lattice Hubbard model at half-filling for various couplings U and temperatures T.
\end{abstract}
Committee: Lucas Wagner, ...
\pagebreak

@@

I propose using a quantum Monte Carlo (QMC) density matrix downfolding (DMD) procedure to develop accurate many-body model Hamiltonians for single (SLG) and bi-layer (BLG) graphene systems.
Recent observations of a fractional quantum Hall effect and plasmarons in SLG as well as unconventional superconductivity in TBLG have spurred an interest in developing many-body models for graphene.
The DMD procedure provides a framework for developing such a many-body model using QMC calculations and starting from the \textit{ab-initio} Hamiltonian 
\begin{equation}
\hat{H}_\text{ab} = -\frac{1}{2} \sum_{i} \nabla_i^2 -\frac{1}{2\mu}\sum_{I} \nabla_I^2 + \frac{1}{2}\sum_{ij} \frac{1}{|r_i - r_j|} + \frac{1}{2}\sum_{IJ} \frac{Z_I Z_J}{|R_I - R_J|} - \sum_{iJ}\frac{Z_J}{|r_i - R_J|},\ \mu = (M/m)
\end{equation}
where $i$ index electrons, $I$ ions, and energy in Hartree (Ha).
The procedure makes no assumptions about the form of microscopic electronic interactions in the system and reports a clear, statistical measure of model quality.
By developing many-body models for SLG and BLG, I hope to further understanding of interactions in graphene like systems and the maturity of DMD for its application to other strongly correlated systems.

\section{Methods}
\subsection{Density Matrix Downfolding (DMD)}
We begin by defining what we mean by a low-energy effective Hamiltonian and downfolding.
Consider a Hamiltonian $\hat{H}$ defined on a Hilbert space $\mathcal{H}$.
A low-energy effective Hamiltonian is an operator $\hat{H}_\text{eff}$ which can accurately reproduce the spectra, up to a constant, and eigenstates of $\hat{H}$ on a subpsace $\mathcal{LE}$ of the Hilbert space $\mathcal{H}$.
Here $\mathcal{LE}$ is taken to be the span of the N lowest energy eigenstates of $\hat{H}$, and $\hat{H}_\text{eff}$ is a linear combination of Hermitian operators $\hat{d}_k$ with a constant energy shift $E_0$
\begin{equation}
\hat{H}_\text{eff} = \sum_{k} g_k \hat{d}_k  + E_0.
\label{eq:Heff}
\end{equation}
The objective of Hamiltonian downfolding is to construct such an $\hat{H}_\text{eff}$ given $\hat{H}, \mathcal{H}$.

The key insight of DMD is that the Hamiltonian downfolding problem can be mapped onto a linear regression problem.
It has been shown that the above definition of $\hat{H}_\text{eff}$ is equivalent to the following: a low-energy effective Hamiltonian is an operator on a subspace $\mathcal{LE}$ of $\mathcal{H}$ such that 
\begin{equation}
\begin{split}
\forall |\Psi\rangle \in \mathcal{LE},\ (E[\Psi] \equiv \langle \Psi|\hat{H} | \Psi \rangle)  = \\ (E_\text{eff}[\Psi] \equiv \langle \Psi | \hat{H}_\text{eff} | \Psi \rangle) + \epsilon[\Psi]
\end{split}
\label{eq:DMD}
\end{equation}
where $\epsilon[\Psi]$ is the error in the effective theory.
Given the general linear form of $\hat{H}_\text{eff}$ seen in \eqref{eq:Heff}, 
the task of building an effective Hamiltonian then reduces to fitting a linear model that minimizes the error $\epsilon[\Psi]$ over $\mathcal{LE}$.
The independent variables (descriptors) in the fitting are expectation values of the operators $d_k[\Psi] \equiv \langle \Psi |\hat{d}_k|\Psi \rangle$ and the target variable is $E[\Psi]$.

This linear regression problem can be tackled in three steps, beginning with sampling states from $\mathcal{LE}$.
Ideally the linear regression would be conducted over the entire space $\mathcal{LE}$. 
In practice, the space is too large, and instead a representative sample set is drawn and used for fitting.
For each sampled state $|\Psi_s\rangle$, the quantities $E[\Psi_s], \{d_k[\Psi_s]\}$ are calculated to be used in the linear regression.
Optimal sampling schemes vary between systems and can significantly reduce the number of samples required to achieve robust statistics.

Generally, however, one cannot sample the true low-energy subspace and low-energy wave functions are sampled from an approximate subspace $\mathcal{LE}^\prime$.
A model fit to an arbitrary subspace of $\mathcal{H}$ will generally fail in describing the low-energy excitations of $\hat{H}$.
Rather, the choice of approximate subspace must be made carefully to ensure the fit model is transferable to the true low-energy space.
We describe in detail later how we selected $\mathcal{LE}^\prime$ for the CuO molecule and the scheme used to draw samples from it.

Next, a set of candidate descriptors which form the effective Hamiltonian are selected.
Descriptors will typically be selected if their values co-vary with the energy over the set of sampled states. 
Knowledge of the low-energy degrees of the freedom from experiment and physical intuition can help restrict the set of candidate descriptors further.
In principle there is no restriction on the form of the candidate operators other than Hermiticity, but second quantized operators are commonly used.

Typically, candidate operators are written in second quantized form, and an appropriate single particle basis to express these operators must be constructed.
An appropriate choice of single particle basis can simplify the representation of the effective theory and typically depends on the system of interest.
Common choices include Bloch waves and molecular orbitals for delocalized basis elements or Wannier and intrinsic atomic orbitals for localized ones.
The selected single particle basis can be validated by ensuring it accurately describes single particle properties among the sampled states like the total number of electrons in the system.

@@

We take advantage of the variational nature of FN-DMC to sample the low-energy states necessary for DMD.
Consider a set of trial wave functions with varying nodal structures.
The final projected $\Psi_\text{DMC}$ for these different trial functions will be the lowest energy states in $\mathcal{H}$ for the given nodal structures.
If the initial trial wave functions are appropriately chosen, the final projected states will be low-energy states within $\Psi_\text{DMC}$.
We calculate the expectation values $E[\Psi_\text{DMC}], \{d_k[\Psi_\text{DMC}]\}$ necessary for the model regression using a mixed estimator.
Details for calculating reduced density matrix elements in FN-DMC can be found in Wagner.

\section{Preliminary Work}
\subsection{Non-orthogonal determinants in FN-DMC trial wave functions}
In order to become acquainted with our code QWalk \cite{WAGNER20093390} and QMC algorithms in general I worked on implementing and testing multi-Slater-Jastrow trial functions with optimized non-orthogonal determinants (MSJ+NO) in FN-DMC \cite{Pathak2018}.
The MSJ+NO trial wave functions take the form
\begin{equation}
\Psi=e^{J(\vec{\alpha})} \sum_I e^{\hat{\Theta}(\vec{\theta_I})} C_I |D_I (\{ \phi\})\rangle
\end{equation}
where the determinants $|D_I\rangle$ are composed of single particle orbitals from a self-consistent field calculation, $e^J$ is a spin-independent Jastrow factor, $e^{\hat{\Theta}}$ is a spin-dependent orbital rotation matrix, and the wave function has variable parameters $\vec{\alpha}, \vec{\theta}, \vec{C}$ which can be optimized.
We assessed the efficiency and compactness of this new trial function by calculating the ground state energy and single particle densities of a C$_2$ molecule using FN-DMC and comparing to the results when using multi-Slater-Jastrow trial functions with optimized orthogonal determinant trial functions (MSJ+O). 
The workflow involved constructing the un-optimized trial wave functions, optimizing the parameters using an energy optimization method \cite{Toulouse2007}, and finally using the optimized trial functions in both variational Monte Carlo (VMC) and FN-DMC calculations.

We found the average improvement in variational energy when using optimized orthogonal determinants is $\langle E_\text{MSJ} - E_\text{MSJ+O} \rangle = $ 0.32 eV with a standard deviation of 0.08 eV, with an additional improvement from the non-orthogonal determinants of 0.031 eV with a standard deviation of 0.021 eV.
The average improvement in the FN-DMC energy with the MSJ+O trial wave functions was 0.14 eV (standard deviation 0.03 eV) with an additional reduction of 0.032 eV (standard deviation 0.019 eV) when using the MSJ+NO trial wave function.
While the benefit using the MSJ+NO trial wave function seems small, we saw that the FN-DMC energy calculated using an MSJ+NO trial function with only 24 determinants was lower than the FN-DMC energy using an MSJ+O trial function with 55 determinants.
Further, the FN-DMC charge density calculated using MSJ+NO trial functions had stronger bonding character than when using MSJ+O trial functions, a reasonable result as introducing correlations into trial functions allows for electrons to avoid each other while still occupying the same bonding region. 
Our results indicated that using non-orthogonal determinants may lead to more compact multi-Slater-Jastrow trial wave functions for small molecules.

\subsection{Effective theory for CuO molecule using DMD}
As a first step towards developing accurate models for extended systems like solids using DMD, I constructed a many-body effective model for the CuO molecule which accurately describes the lowest-lying eigenstates seen in experiment.
The neutral CuO molecule has been a subject of intense theoretical and experimental study due to the complex structure of its low-energy space.
Unlike lighter transition metal oxide molecules like TiO or FeO, the energy scales related to the active Cu 3d, 4s and O 2p orbitals are very similar.
This leads to many low-energy eigenstates below $\sim $ 3eV which have radically different properties.
As such, no accurate low-energy effective model for the molecule has been constructed successfully.

In the following four sections I will detail the steps I took towards developing a model for the CuO molecule using DMD.
First, I discuss how low-energy states were sampled for the model regression.
This is followed by details of descriptor selection and single particle basis construction.
With the sampled data and candidate descriptors in hand, the feature selection and model fitting procedures are described.
I will conclude by comparing the solutions of the model Hamiltonian to experiment and previous theoretical calculations.

\subsubsection{Sampling low-energy states}
For the CuO molecule, low-energy states were generated by using FN-DMC projected multi-Slater-Jastrow trial wave functions.
The trial wave functions took the form
\begin{equation}
|\Psi_T(\vec{p}) \rangle =  e^{J}\sum_{j} p_j|\text{D}_j\rangle,
\label{eq:sampling}
\end{equation}
where $e^J$ is a Jastrow factor, $|\text{D}_j\rangle$ are selected determinants which we discuss below, and $p_j$ are variable coefficients.
A single three-body Jastrow factor was used for all sampled states optimized on the lowest energy determinant $|\text{D}_0 \rangle$.
The choice to use a single Jastrow factor significantly reduced the computational cost of the DMD method and is justified by noting that large variations in the Jastrow factor would move the state outside the low-energy space.

The determinants $|\text{D}_j \rangle$ were constructed from the orbitals of independent symmetry-targeted unrestricted Kohn Sham (UKS) calculations.
This choice was made to ensure that the three major low-energy variations seen in the molecule can be captured in our samples: occupancy, relaxation and spin configuration of the Cu 3d, 4s and O 2p orbitals.
The symmetry-targeting allows us to select the number of electrons in both the spin-up and spin-down channel for a given irreducible representation of the symmetry group for CuO, letting us generate determinants with variations in both total spin and orbital occupation.
Orbital relaxation effects are captured by having each determinant, corresponding to a unique orbital and spin configuration, be the result of an independent self-consistent field calculation.
The UKS calculations were done using PySCF, a B3LYP functional, Trail-Needs psuedopotential and VTZ Trail-Needs basis.

Of these UKS determinants, only those which contribute to accurately describing the low-lying excitations of the molecule are selected.
All determinants with at most a single hole in the Cu 3d shell and no doubly occupied Cu 4s orbital were included in \eqref{eq:sampling} to reflect the low-energy excitations observed in experiment.
Only the lowest energy state with a Cu 4s$^2$ electronic configuration was included, corresponding to a double excitation 2 O ${2p_\pi} \rightarrow$ 2 Cu 4s with an energy $>7$ eV above the UKS ground state.
Inclusion of this high energy determinant is necessary to ensure our model places similar Cu 4s$^2$ states high in energy.
No other Cu 4s$^2$ determinants were included since we were not interested in describing the variation among these high energy states.

A shell sampling scheme was used to sample states from the selected approximate low-energy space.
We began by fixing the coefficient $p_0 = \sqrt{w}$ where $w \in \{1.0, 0.8, ..., 0.2\}$. 
For each choice of $w$ we sampled n = 5 states by randomly selecting the unassigned coefficients from independent, standard normal distributions such that $\sum_j p_j^2 = 1$. 
This procedure generated a uniform set of samples from spherical shells at decreasing radii in determinant space centered at $|\text{D}_0\rangle$.
We repeated this process centered at every determinant in \eqref{eq:sampling} resulting in a sample set of overlapping shells which fill the space $\mathcal{LE}^\prime$.

The final sampled states were projected using FN-DMC to ensure they are as low in energy as possible.
The FN-DMC calculations were conducted with T-moves to ensure the method is variational while using pseudopotentials.
A timestep of $\tau = 0.01$ was used.

\subsubsection{Selecting candidate descriptors}
We select the following set of 1- and 2-body operators to construct our candidate descriptors based on our understanding of the low-energy excitations of the CuO molecule and our sampled states:
\begin{equation}
\begin{split}
\sum_{\sigma = \uparrow, \downarrow} \Bigg(\epsilon_{3d_\pi}\hat{n}_{3d_\pi,\sigma} + \epsilon_{3d_{z^2}}\hat{n}_{3d_{z^2},\sigma} +  \epsilon_{2p_z} \hat{n}_{2p_z,\sigma} + \epsilon_{2p_\pi}\hat{n}_{2p_\pi,\sigma} + \epsilon_{4s}\hat{n}_{4s,\sigma} +\\
t_\pi \hat{c}_{3d_\pi,\sigma}^\dagger \hat{c}_{2p_\pi,\sigma} + t_{dz} \hat{c}_{3d_{z^2},\sigma}^\dagger \hat{c}_{2p_z,\sigma}  + t_{sz}\hat{c}_{4s,\sigma}^\dagger \hat{c}_{2p_z,\sigma} + t_{sd}\hat{c}_{4s,\sigma}^\dagger \hat{c}_{3d_{z^2},\sigma} + \text{h.c.} + \Bigg)  \\
J_{sd}\sum_{i\in {\{\text{xy, xz, ...}}\}} \vec{S}_{4s} \cdot \vec{S}_{3d_i} + U_s \hat{n}_{4s,\uparrow}\hat{n}_{4s,\downarrow} + E_0.
\end{split}
\label{eq:models}
\end{equation}
The operator subscripts correspond to the intrinsic atomic orbital (IAO) basis elements constructed using the molecular orbitals in \eqref{eq:sampling}, and can be seen in \ref{fig:cuo_regr}a.
Operators are contracted over symmetry-related pairs for the $3d_\pi, 2p_\pi$ orbitals, for example $\hat{n}_{3d_\pi} = \hat{n}_{3d_{xz}} + \hat{n}_{3d_{yz}}$.

@@

The operator $\hat{n}_{3d_\delta}$ is not included as it is linearly dependent on the rest of the occupation operators.
The linear dependency arises from the fixed number of electrons in the model.
As such, the operator $\hat{n}_{3d_\delta} = N_\text{elec} - \sum_{i \neq 3d_\delta} \hat{n}_i$.
The first term can be absorbed into the constant $E_0$, and the latter terms shift the occupation energies $\epsilon_i$.

\subsubsection{Fitting the model}
A direct attempt at fitting all eleven terms in \eqref{eq:models} leads to a high goodness of fit but unexpectedly large coefficients for the fit model.
The fit model had an R$^2= 0.99$, but the fit coefficients were very inconsistent with experimental measurements and our sampled data.
For example, the model predicted an enormous occupation energy gap of $\epsilon_{2p_\pi} - \epsilon_{3d_\delta} = $ 5.87 eV, while in experiment and our sampled data this scale was between 2 $-$ 3 eV.
This inflation of model parameters was found to be associated with multicollinearity among the candidate descriptors, quantified by a large condition number (CN) of 1470.

A principal component analysis (PCR) analysis was used to alleviate this multicollinearity and fit the effective theory.
PCR can be broken into three parts: conducting a principal component analysis (PCA) and selecting candidate principal components to work with (PCs), fitting a linear model to these PCs, and then inverting the coefficients back to the original descriptors.
If the subset of PCs are chosen appropriately, the final regressed model will have a similar goodness of fit to one fit on the original data matrix but without the consequence of serious multicollinearity.

From the PCA, the nine highest ranked PCs were selected to use in the PCR.
In Figure \eqref{fig:cuo_regr}b we plot both the cumulative explained variance ratio (CEVR) and CN of a matrix composed of the $N_\text{comp}$ highest ranked PCs varying from one to eleven.
The CEVR indicates how much of the variation in the descriptor values a subset of PCs capture, and the CN quantifies the multicollinearity among these PCs.
The objective in selecting PCs is to the maximize the prior while simultaneously minimizing the latter, ensuring a high goodness of fit with little to no multicollinearity.
Reducing the number of PCs from eleven rapidly decreases the CN to $\sim $ 200 while keeping the CEVR relatively fixed.
Further reducing the number of PCs past nine does not affect the CN but steadily decreases the CEVR, indicating that nine principal components best satisfies the objective above with a CEVR 0.9995 and CN of 257.

The model fit to these nine PCs maintains an excellent goodness of fit without serious multicollinearity.
When trained on all the sampled data this model yields an $R^2$ = 0.98 with a CN of 257. 
While the goodness of fit does decrease slightly compared to a fit with the full data matrix, the nearly six-fold reduced condition number readily compensates for this loss.
An inverse PCA transformation was used to map the fit coefficients corresponding to the PC descriptors back into the original eleven coefficients of \eqref{eq:models}.
These coefficients can be seen in Figure \ref{fig:cuo_regr}c with single standard deviation bootstrap error estimates.

>>>>

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{subfigure}
\usepackage[margin=1.0in]{geometry}
\usepackage{float}
\usepackage{ulem}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{wrapfig}
\renewcommand{\baselinestretch}{1.5}
\usepackage[utf8]{inputenc}
\author{Shivesh Pathak}
\title{Developing effective theories for single and bilayer graphene using density matrix downfolding}

\begin{document}
\maketitle
\begin{abstract}
A core problem of condensed matter physics is understanding how to develop low-energy effective models for realistic systems.
I believe that a density matrix downfolding (DMD) method can be used to systematically build effective theories for such systems beginning with a high energy theory like the \textit{ab-initio} Hamiltonian.
I present as a preliminary result an effective Hamiltonian developed using DMD which can accurately reproduce the spectra and properties of the lowest lying excitations of the CuO molecule.
I propose using DMD to develop accurate model Hamiltonians for single and bilayer graphene.
My work will serve to mature the DMD method and take steps towards a model for the strongly correlated twisted bilayer graphene system.
\end{abstract}
Committee: Lucas Wagner, Smitha Vishveshwara, David Ceperley, Karin Dahmen

Pool: \textit{David Ceperley},\textit{ Brian DeMarco}, Smitha Vishveshwara, Andre Schleife, \textit{Karin Dahmen} 
\pagebreak

@@

I propose using a quantum Monte Carlo (QMC) density matrix downfolding (DMD) procedure to develop accurate many-body model Hamiltonians for single (SLG) and bi-layer (BLG) graphene systems.
Recent observations of a fractional quantum Hall effect and plasmarons in SLG as well as unconventional superconductivity in TBLG have spurred an interest in developing many-body models for graphene.
The DMD procedure provides a framework for developing such a many-body model using QMC calculations starting from the \textit{ab-initio} Hamiltonian 
\begin{equation}
\hat{H}_\text{ab} = -\frac{1}{2} \sum_{i} \nabla_i^2 -\frac{1}{2\mu}\sum_{I} \nabla_I^2 + \frac{1}{2}\sum_{ij} \frac{1}{|r_i - r_j|} + \frac{1}{2}\sum_{IJ} \frac{Z_I Z_J}{|R_I - R_J|} - \sum_{iJ}\frac{Z_J}{|r_i - R_J|},\ \mu = (M/m)
\label{Hab}
\end{equation}
where $i$ index electrons, $I$ ions, and energy in Hartree (Ha).
The procedure makes no assumptions about the form of microscopic electronic interactions in the system and reports a clear, statistical measure of model quality.
By developing many-body models for SLG and BLG, I hope to further the understanding of interactions in graphene like systems and the maturity of DMD for its application to other strongly correlated systems like TBLG.

\subsection{Density Matrix Downfolding (DMD)}
I begin by defining what a low-energy effective Hamiltonian and downfolding are.
Consider a Hamiltonian $\hat{H}$ defined on a Hilbert space $\mathcal{H}$.
A low-energy effective Hamiltonian is an operator $\hat{H}_\text{eff}$ which can accurately reproduce the spectra, up to a constant, and eigenstates of $\hat{H}$ on a subpsace $\mathcal{LE}$ of the Hilbert space $\mathcal{H}$.
Here $\mathcal{LE}$ is taken to be the span of the N lowest energy eigenstates of $\hat{H}$, and $\hat{H}_\text{eff}$ is a linear combination of Hermitian operators $\hat{d}_k$ with a constant energy shift $E_0$
\begin{equation}
\hat{H}_\text{eff} = \sum_{k} g_k \hat{d}_k  + E_0.
\label{eq:Heff}
\end{equation}
The objective of Hamiltonian downfolding is to construct such an $\hat{H}_\text{eff}$ given $\hat{H}, \mathcal{H}$.

The key insight of DMD is that the Hamiltonian downfolding problem can be mapped onto a linear regression problem.
It has been shown that the above definition of $\hat{H}_\text{eff}$ is equivalent to the following: a low-energy effective Hamiltonian is an operator on a subspace $\mathcal{LE}$ of $\mathcal{H}$ such that 
\begin{equation}
\begin{split}
\forall |\Psi\rangle \in \mathcal{LE},\ (E[\Psi] \equiv \langle \Psi|\hat{H} | \Psi \rangle)  = (E_\text{eff}[\Psi] \equiv \langle \Psi | \hat{H}_\text{eff} | \Psi \rangle) + \epsilon[\Psi]
\end{split}
\label{eq:DMD}
\end{equation}
where $\epsilon[\Psi]$ is the error in the effective theory.
Given the general linear form of $\hat{H}_\text{eff}$ seen in \eqref{eq:Heff}, 
the task of building an effective Hamiltonian then reduces to fitting a linear model that minimizes the error $\epsilon[\Psi]$ over $\mathcal{LE}$.
The independent variables (descriptors) in the fitting are expectation values of the operators $d_k[\Psi] \equiv \langle \Psi |\hat{d}_k|\Psi \rangle$ and the target variable is $E[\Psi]$.

This linear regression problem can be tackled in three steps, beginning with sampling states from $\mathcal{LE}$.
Ideally the linear regression would be conducted over the entire space $\mathcal{LE}$. 
In practice, the space is too large, and instead a representative sample set is drawn and used for fitting.
Generally, one cannot sample the true low-energy subspace, and low-energy wave functions are sampled from an approximate subspace $\mathcal{LE}^\prime$.
For each sampled state $|\Psi_s\rangle$, the quantities $E[\Psi_s], \{d_k[\Psi_s]\}$ are calculated to be used in the linear regression.
The sampling and calculation are typically done in QMC, but less accurate methods like DFT and Hartree-Fock can be used as well, resulting is less accurate models.

Next, a set of candidate descriptors which form the effective Hamiltonian are selected.
Descriptors will typically be selected if their values co-vary with the energy over the set of sampled states. 
Knowledge of the low-energy degrees of the freedom from experiment and physical intuition can help restrict the set of candidate descriptors further.
In principle there is no restriction on the form of the candidate operators other than Hermiticity, but second quantized operators are commonly used.
If so, an appropriate single particle basis to express the second quantized operators must be constructed.

@@

One can take advantage of the variational nature of FN-DMC to sample the low-energy states necessary for DMD.
Consider a set of trial wave functions with varying nodal structures.
The final projected $\Psi_\text{DMC}$ for these different trial functions will be the lowest energy states in $\mathcal{H}$ for the given nodal structures.
If the initial trial wave functions are appropriately chosen, the final projected states will be low-energy states within $\Psi_\text{DMC}$.
The expectation values $E[\Psi_\text{DMC}], \{d_k[\Psi_\text{DMC}]\}$ necessary for the model regression are calculated using a mixed estimator.
Details for calculating reduced density matrix elements in FN-DMC can be found in Wagner.

\section{Preliminary Work}
\subsection{Non-orthogonal determinants in FN-DMC trial wave functions}
In order to become acquainted with our code QWalk \cite{WAGNER20093390} and QMC algorithms in general I worked on implementing and testing multi-Slater-Jastrow trial functions with optimized non-orthogonal determinants (MSJ+NO) in FN-DMC \cite{Pathak2018}.
The MSJ+NO trial wave functions take the form
\begin{equation}
\Psi=e^{J(\vec{\alpha})} \sum_I e^{\hat{\Theta}(\vec{\theta_I})} C_I |D_I (\{ \phi\})\rangle
\end{equation}
where the determinants $|D_I\rangle$ are composed of single particle orbitals from a self-consistent field calculation, $e^J$ is a spin-independent Jastrow factor, $e^{\hat{\Theta}}$ is a spin-dependent orbital rotation matrix, and the wave function has variable parameters $\vec{\alpha}, \vec{\theta}, \vec{C}$ which can be optimized.

We assessed the efficiency and compactness of this new trial function by calculating the ground state energy and single particle densities of a C$_2$ molecule using FN-DMC and comparing to the results when using multi-Slater-Jastrow trial functions with optimized orthogonal determinant trial functions (MSJ+O). 
The workflow involved constructing the un-optimized trial wave functions, optimizing the parameters using an energy optimization method \cite{Toulouse2007}, and finally using the optimized trial functions in both variational Monte Carlo (VMC) and FN-DMC calculations. 
We found the average improvement in variational energy when using optimized orthogonal determinants is $\langle E_\text{MSJ} - E_\text{MSJ+O} \rangle = $ 0.32 eV with a standard deviation of 0.08 eV, with an additional improvement from the non-orthogonal determinants of 0.031 eV with a standard deviation of 0.021 eV.
The average improvement in the FN-DMC energy with the MSJ+O trial wave functions was 0.14 eV (standard deviation 0.03 eV) with an additional reduction of 0.032 eV (standard deviation 0.019 eV) when using the MSJ+NO trial wave function.

While the benefit using the MSJ+NO trial wave function seems small, we saw that the FN-DMC energy calculated using an MSJ+NO trial function with only 24 determinants was lower than the FN-DMC energy using an MSJ+O trial function with 55 determinants.
Further, the FN-DMC charge density calculated using MSJ+NO trial functions had stronger bonding character than when using MSJ+O trial functions, a reasonable result as introducing correlations into trial functions allows for electrons to avoid each other while still occupying the same bonding region. 
Our results indicated that using non-orthogonal determinants may lead to more compact multi-Slater-Jastrow trial wave functions for small molecules.

\subsection{Effective theory for CuO molecule using DMD}
As a first step towards developing accurate models for extended systems like solids using DMD, I constructed a many-body effective model for the CuO molecule by downfolding \eqref{Hab} under a Born-Oppenheimer approximation with fixed bond length $r_e$ = 1.725\r{A}.
In the following four sections I will detail the steps I took towards developing a model for the CuO molecule using DMD.
First, I discuss how low-energy states were sampled for the model regression.
This is followed by details of descriptor selection and single particle basis construction.
With the sampled data and candidate descriptors in hand, the feature selection and model fitting procedures are described.
I will conclude by comparing the solutions of the model Hamiltonian to experiment and previous theoretical calculations.

\paragraph{Sampling low-energy states}
For the CuO molecule, low-energy states were generated by using FN-DMC projected multi-Slater-Jastrow (MSJ) trial wave functions.
The trial wave functions took the form
\begin{equation}
|\Psi_T(\vec{p}) \rangle =  e^{J}\sum_{j} p_j|\text{D}_j\rangle,
\label{eq:sampling}
\end{equation}
where $e^J$ is a Jastrow factor, $|\text{D}_j\rangle$ are selected determinants which we discuss below, and $p_j$ are variable coefficients.
A single three-body Jastrow factor was used for all sampled states optimized on the lowest energy determinant $|\text{D}_0 \rangle$.
This significantly reduced the computational cost of the DMD method and is justified by noting that large variations in the Jastrow factor would move the state outside the low-energy space.

The determinants $|\text{D}_j \rangle$ were constructed from the orbitals of independent symmetry-targeted unrestricted Kohn Sham (UKS) calculations.
This choice was made to ensure that the three major low-energy variations seen in the molecule can be captured in our samples: occupancy, spin configuration and relaxation of the Cu 3d, 4s and O 2p orbitals.
All determinants with at most a single hole in the Cu 3d shell and no doubly occupied Cu 4s orbital were included in \eqref{eq:sampling} to reflect the low-energy excitations seen in experiment.
Only the lowest energy state with a Cu 4s$^2$ electronic configuration was included, corresponding to a double excitation 2 O ${2p_\pi} \rightarrow$ 2 Cu 4s with an energy $>7$ eV above the UKS ground state.
Inclusion of this high energy determinant is necessary so our model places similar Cu 4s$^2$ states high in energy.
The UKS calculations were done using PySCF, a B3LYP functional, Trail-Needs psuedopotential and VTZ Trail-Needs basis.

A shell sampling scheme was used to sample states from the selected approximate low-energy space.
We began by fixing the coefficient $p_0 = \sqrt{w}$ where $w \in \{1.0, 0.8, ..., 0.2\}$. 
For each choice of $w$ we sampled n = 5 states by randomly selecting the unassigned coefficients from independent, standard normal distributions such that $\sum_j p_j^2 = 1$. 
This procedure generated a uniform set of samples from spherical shells at decreasing radii in determinant space centered at $|\text{D}_0\rangle$.
We repeated this process centered at every determinant in \eqref{eq:sampling} resulting in a sample set of overlapping shells which fill $\mathcal{LE}^\prime$.

The final sampled MSJ states were projected using FN-DMC to generate our low-energy states.
The FN-DMC calculations were conducted with T-moves to ensure the variational principle holds while using pseudopotentials.
A timestep of $\tau = 0.01$ was used.

\paragraph{Selecting candidate descriptors}
We selected the following set of 1- and 2-body operators to construct our candidate descriptors based on our understanding of the low-energy excitations of the CuO molecule and our sampled states:
\begin{equation}
\begin{split}
\sum_{\sigma = \uparrow, \downarrow} \Bigg(\epsilon_{3d_\pi}\hat{n}_{3d_\pi,\sigma} + \epsilon_{3d_{z^2}}\hat{n}_{3d_{z^2},\sigma} +  \epsilon_{2p_z} \hat{n}_{2p_z,\sigma} + \epsilon_{2p_\pi}\hat{n}_{2p_\pi,\sigma} + \epsilon_{4s}\hat{n}_{4s,\sigma} +\\
t_\pi \hat{c}_{3d_\pi,\sigma}^\dagger \hat{c}_{2p_\pi,\sigma} + t_{dz} \hat{c}_{3d_{z^2},\sigma}^\dagger \hat{c}_{2p_z,\sigma}  + t_{sz}\hat{c}_{4s,\sigma}^\dagger \hat{c}_{2p_z,\sigma} + t_{sd}\hat{c}_{4s,\sigma}^\dagger \hat{c}_{3d_{z^2},\sigma} + \text{h.c.} + \Bigg)  \\
J_{sd}\sum_{i\in {\{\text{xy, xz, ...}}\}} \vec{S}_{4s} \cdot \vec{S}_{3d_i} + U_s \hat{n}_{4s,\uparrow}\hat{n}_{4s,\downarrow} + E_0.
\end{split}
\label{eq:models}
\end{equation}
The operator subscripts correspond to the intrinsic atomic orbital (IAO) basis elements constructed using the molecular orbitals in \eqref{eq:sampling}, and can be seen in \ref{fig:cuo_regr}a.

@@

\paragraph{Fitting the model}
A direct attempt at fitting all eleven terms in \eqref{eq:models} leads to a high goodness of fit, R$^2$ = 0.99, but unexpectedly large coefficients for the fit model, like an enormous occupation energy gap of $\epsilon_{2p_\pi} - \epsilon_{3d_\delta} = $ 5.87 eV, which is 2 $-$ 3 eV in experiment and our sampled data.
This inflation of model parameters was found to be associated with multicollinearity among the candidate descriptors, quantified by a large condition number (CN) of 1470.

A principal component regression (PCR) was used to alleviate this multicollinearity and fit the effective theory.
PCR can be broken into three parts: conducting a principal component analysis (PCA) and selecting candidate principal components to work with (PCs), fitting a linear model to these PCs, and then inverting the coefficients back to the original descriptors.
If the subset of PCs are chosen appropriately, the final regressed model will have a similar goodness of fit to one fit on the original data matrix but without the consequence of serious multicollinearity.

From the PCA, the nine highest ranked PCs were selected to use in the PCR.
In Figure \eqref{fig:cuo_regr}b we plot both the cumulative explained variance ratio (CEVR) and CN of a matrix composed of the $N_\text{comp}$ highest ranked PCs varying from one to eleven.
The CEVR indicates how much of the variation in the descriptor values a subset of PCs capture, and the CN quantifies the multicollinearity among these PCs.
The objective in selecting PCs is to the maximize the prior while simultaneously minimizing the latter, ensuring a high goodness of fit with little to no multicollinearity.
Nine principal components best satisfies the objective above with a CEVR 0.9995 and CN of 257.

The model fit to these nine PCs maintains an excellent goodness of fit without serious multicollinearity.
When trained on all the sampled data this model yields an $R^2$ = 0.98 with a CN of 257. 
An inverse PCA transformation was used to map the fit coefficients corresponding to the PC descriptors back into the original eleven coefficients of \eqref{eq:models}.
These coefficients can be seen in Figure \ref{fig:cuo_regr}c with single standard deviation bootstrap error estimates.

\begin{figure*}
\centering
\hspace*{-0.25in}
\includegraphics[width=1.1\linewidth]{./figs/cuo_regr_2.pdf}
\caption{Figures related to model regression for the CuO molecule: a) Intrinsic atomic orbitals used as a single particle basis for the model Hamiltonian, b) Results of PCA showing the CEVR and CN of a data set containing N$_{comp}$  highest ranked PCs, c) The final regressed parameters from the PCR, d) Comparison of four lowest lying eigenstates of CuO molecule between experiment and three computation methods.}
\label{fig:cuo_regr}
\end{figure*}

\paragraph{Solutions of the model}
The model is solved by exact diagonalization, and the solutions agree well with experimental spectra and state assignments.
Shown in Figure \ref{fig:cuo_regr}d is a comparison between the excitation energies calculated from our model and those seen in the latest anion photoelectron spectroscopy measurements.
The model energies for these eigenstates are denoted by orange circles and agree with experiment up to a maximum deviation of 0.2 eV.
The term symbols assigned in experiment also match the term symbols of the model eigenstates, which are constructed by looking at the multiplicity and electronic configuration of each model eigenstate.
